Diversification of Australian blind snakes

### 2022-11-24 manuscript.Rmd
add results from GeoHiSSE

To investigate how the distribution of biome evolved in blind snakes we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including likelihood implementations of Dispersal-Vicariance Analysis [@ronquistDispersalVicariance1997] (DIVALIKE), Dispersal-Extinction-Cladogenesis [@reeMaximum2008] (DECLIKE), and BayArea [@landisBayesian2013] (BAYAREALIKE) in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought is relevant. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). 
We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the model that fitted the data best based on corrected Akaike Information Criterion (AICc) scores. Based on the best model we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and trasition between biomes, and to estimate the frequency and directionality of dispersal events across biomes [@mat.

@@

We further investigated whether diversification rates and biome transitions differ between geographic areas. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction (SSE) models under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the phylogeny and geographic range data. 
Since SSE models only allow two states, we restricted our geographic states to species that have distributions in arid areas, non-arid areas, and those that are widespread across both arid and non-arid areas. 
Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five SSE models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved 'hidden states'. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

@@

We identified the DEC+_j_+_x_ model as the most preferred biogeographical history model (delta AICc = -4.47, see Table-supp-BioGeoBEARS). 
Likelihood ratio test (LRT) indicates that both founder events (+*j* parameter) and distance between biome (+x parameter) together provides statistically significant improvement over the simpler DEC model, DEC+x  and DEC+j (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001). 
Our best model inferred tropical grassland biome to be the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears)), but the origin of crown *Anilios* is ambiguous.

Following the hypothesized colonisation event from a single common ancestor, the diversifcation of _Anilios_ occurred very rapidly. Descendents of the earliest split remained in the tropical grassland biome but within a million years later another split occured. The earliest biome transition was from tropical grassland biome into the temperate forest biome around 25 MYA (95% HPD: 28 – 22 MYA) for branch leading _A. siliva_, and another branch that included current arid and mediterranean species starting 24 MYA (95% HPD: 28 – 21 MYA). Based on 50 biogeographic stochastic mapping, we found that the largest source of total outward dispersal (anagenetic and cladogenetic) is the tropical grassland biome (mean = XX) almost double the number of dispersals from the arid biome (mean = XX) (\@ref(fig:fig-biogeobears)). On the other hand, the largest sink for all inward dispersal is the mediterranean biome (

The top recipients of dispersals from tropical grassland biome is the temperate grassland (5.06 ± 0.79), follotropical forest (4.12 ± 1.29), ,

Subsequent splits show transitions from the arid biome into the mediterranean biome starting after 15 MYA (\@ref(fig:fig-biogeobears). Current distributions of _Anilios_ are mostly concentrated in the tropical grassland biome closely followed by the Arid zone and Mediterranean biome. We observed that closely related species tend to share the same biome and occur around the same geographical areas.

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods \@ref(fig:fig-paleoclimate-compare)b. However, all three paleo temperature models indicates that Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant compared to global temperature that shows greater fluctuation then slight decrease towards the present.

@@

Rates of diversification between arid and non-arid, and widespread.

>>>>

To investigate how the distribution of biome evolved in blind snakes we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including likelihood implementations of Dispersal-Vicariance Analysis [@ronquistDispersalVicariance1997] (DIVALIKE), Dispersal-Extinction-Cladogenesis [@reeMaximum2008] (DECLIKE), and BayArea [@landisBayesian2013] (BAYAREALIKE) in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought is relevant. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). 
We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the model that fitted the data best based on corrected Akaike Information Criterion (AICc) scores. Based on the best model we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and trasition between biomes, and to estimate the frequency and directionality of dispersal events across biomes [@matzkeBioGeoBEARS2013].

We further investigated whether diversification rates and biome transitions differ between arid and non-arid biomes. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction (SSE) models under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the phylogeny and geographic range data. 
Since SSE models only allow two states, we restricted our geographic states to species that have distributions in arid areas, non-arid areas, and those that are widespread across both arid and non-arid areas. 
Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five SSE models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved 'hidden states.' Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivations of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

@@

@@

<!-- ## values and info from 04_BioGeoBEARS_bsm_plots.R -->
We found that _Anilios_ species are mostly found solely or partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and mediterranean biome (n = 10). Closely related species tend to share the same biome (\@ref(fig:fig-biogeobears). 
We identified the DEC+_j_+_x_ model as the most preferred biogeographical history model (delta AICc = -4.47, see Table-supp-BioGeoBEARS). 
Likelihood ratio test (LRT) indicates that both founder events (+*j* parameter) and distance between biome (+x parameter) together provides statistically significant improvement over the simpler DEC model, DEC+x  and DEC+j (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001).

Based on 50 BioGeoBEARS biogeographic stochastic mapping of our best model, we inferred the tropical grassland biome to be the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears)), but the origin of crown *Anilios* is ambiguous. We estimated an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occured throughout the phylogeny, out of which the majority were a combination of anagenetic dispersal (31.2 ± 1.3 events) and founder events (6.1 ± 1.5 events) (see supp-summary_counts_bsm_table). 
We found that the largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events; 43.6%) making up almost double the number of outward dispersals from the arid biome (9.7 ± 3.2 events; 26.0%) (\@ref(fig:fig-biogeobears))c-d. Together the tropical grassland biome and the arid biome make up the majority (69.6%) of outward transitions.

On the other hand, the largest sink for all inward dispersal are the temperate grassland biome (7.9 ± 4.2 events; 21.2%) and the mediterranean biome (7.0 ± 2.9 events; 18.8%). The largest directional transition is from the arid biome to the mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (\@ref(fig:fig-biogeobears)c-d.

We applied GeoHiSSE models to evaluate whether speciation, extinction, and turnover rates differ between lineages in arid, non-arid, and widespread ranges. We found that the model where where geographic range was not linked to variation in diversification rates but rather to a hidden unobserved state fit the data better than models where diversification rates were depdendent on geographic range (see supp-table-geohisse). The best-fitting model accounts for 0.95 of the AIC weight and had AIC value 7.02 units lower than that of the second best-fitting model (see suupp-table-geohiss). The weighted average parameter estimates indicate both diversification rates and turnover rates did not differ between arid and non-arid ranges (REF FIGURE).

Results from our ClaDS analysis is consistent with that of GeoHiSSE, showing branch-specific diversification rates slow down towards the present.

The analysis shows slow down in diversification rates over time.

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods (\@ref(fig:fig-paleoclimate-compare))b. However, all three paleo temperature models indicates that Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant compared to global temperature that shows greater fluctuation then slight decrease towards the present.

@@
